---
title: "Class 11"
author: "Cynthia Perez (A16393492)"
format: pdf
---

# Section 1. Proportion of G/G in a population

Downloaded the csv file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39827525-39962692;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel >

Here we read this CSV file
```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")

head(mxl)
```

We want to find proportion that are homozygous (G|G) for asthma. Use the `table()` function. 
```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) *100
```

# Section 4 Population Scale Analysis

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
#need to read in the file first
expr <- read.table("rs8067378_ENSG00000172057.6.txt", row.names=1)
head(expr)
```

Sample size for each genotype
```{r}
table(expr$geno)
```
Median Expression levels for each genotype
```{r}
medis <- tapply(expr$exp, expr$geno, median)
medis
```



> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? 

Make a boxplot
```{r}
library (ggplot2)
```

```{r}
genobox <- ggplot(expr) + aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
genobox
```
G/G has a lower expression level relative to A/A genotype
The SNP does effect the expression of ORMDL3. 
