---
title: "Class 18: Mini Project - Investigating Pertussis Resurgence"
author: "Cynthia Perez (A16393492)"
format: pdf
---

Pertussis (whopping cough) is a highly contagious lung infection that is most severe in the very young (under 1 year of age)

# Investigating Pertussis Cases by Year

We start with looking at Pertussis case numbers per year in the United States from the CDC website.

> Q1. With the help of the R “addin” package `datapasta` assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

First install datapasta package in console using `install.packages()`
Go to https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html to copy table data onto clipboard
Then use Paste as data.frame to transfer the data from the clipboard into a dataframe.
```{r}
cdc <- data.frame(
                          Year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L,2020L,2021L),
  Reported.Cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617,6124,
                                   2116)
)
```

Using ggplot to make a plot of cases over time
```{r}
library(ggplot2)

casesplot <- ggplot (cdc) +
  aes(x= Year, y= Reported.Cases) +
  geom_point() +
  geom_line() +
  labs(title= "Pertussis Cases by Year (1992-2021)", x="Year", y= "Number of cases")
```


# Two Vaccines (wP and aP)

> Q2. Using the ggplot `geom_vline()` function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine.

```{r}
casesplot + 
  geom_vline(xintercept = 1946, col= "blue", linetype="longdash") + 
  geom_vline(xintercept = 1996, col= "red", linetype="longdash")
```
> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine number of cases reported stayed low for a few years. However small resurgences of cases were observed after the year 2000. This could be due to improved testing sensitivity, hesitancy from parents leading to less vaccination in children, or less effectiveness from the aP vaccine. 

# Exploring CMI-PB Data

A systems vaccine project to figure out what is going on with aP vs wP immune responses. 
Evaluation of pertussis-specific immune responses over time in individuals with wP and aP vaccines. Look into the data from CMI-PB project using jsonlite package, will simplify JSON "key:value" pair arrays into R data frames
 
```{r}
# load in the jsonlite package
library(jsonlite)

# read in main subject database table from the CMI-PB API
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE)
head(subject, 3)
```
 
 > Q4. How many aP and wP infancy vaccinated subjects are in the dataset?
 
```{r}
table(subject$infancy_vac)
```
 > Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```
> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

### Working with dates
Converting year-of-birth and date-of-boost columns from dates into years. We use the `lubridate` package in R

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
#install lubridate package in R console
library(lubridate)

#create a column "age" in subject dataframe for the age of all subjects in years
subject$age <- time_length(today() -ymd(subject$year_of_birth), "years")
```

```{r}
#load up dplyr to filter subjects age by aP vs wP
library(dplyr)

#filter for aP
ap <- subject %>% filter(infancy_vac == "aP")
round(summary(ap$age))

#filter for wP
wp <- subject %>% filter(infancy_vac == "wP")
round(summary(wp$age))

```

### Joining Multiple Tables

Read other tables from the CMi-PB resource
```{r}
specimen <- read_json("http://cmi-pb.org/api/specimen", simplifyVector = T)
ab_titer <- read_json("http://cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector = T)
```

```{r}
head(specimen)
```
```{r}
head(ab_titer)
```
We must link the specimen and subject data frames to determine whether the specimen_id comes from aP or wP individuals. We shall link the tables into a data frame called "meta"

> Q9. join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- inner_join(specimen, subject)

dim(meta)
head(meta)
```

> Q10. Now using the same procedure join meta with "ab_titer" data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(ab_titer, meta)
dim(abdata)
head(abdata)
```
Now we have our merged dataset with all the needed metadata and antibody measurements called "abdata"

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```
> Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

# Examine IgG Ab Titer Levels

Use the dplyr package to again filter our abdata merged dataset by IgG isotype
```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

>Q13. Make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(igg$MFI_normalised, antigen) +
  geom_boxplot() +
  facet_wrap(vars(visit), nrow=2)
```

```{r}
table(igg$visit)
```

Let's dig in a little bit more...

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```

Let's look into the 2021 dataset IgG PT antigen level time-course

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```
